Modern Statistical Methods --- Example Sheet 3

### 2020-11-23 distribution_theory/C4.tex
wtfff

We start with a lemma:
\begin{lemma}
	For $u \in \EE'(\RR^n)$, the Fourier transform $\hat u \in \SS'(\RR^n)$ can be identified with the smooth (analytic) function $\lambda \mapsto \ang{u(x), e^{-i\lambda\vdot x}}$, which we will denote $\hat u(\lambda)$. 
\end{lemma}

@@

>>>>

We start with a lemma:
\begin{lemma}
	For $u \in \EE'(\RR^n)$, the Fourier transform $\hat u \in \SS'(\RR^n)$ can be identified with the smooth (real-analytic) function $\lambda \mapsto \ang{u(x), e^{-i\lambda\vdot x}}$, which we will denote $\hat u(\lambda)$. 
\end{lemma}

@@

\subsection{Paley-Wiener-Schwartz theorem}
We have seen that if $u \in \EE'(\RR^n)$, then $\hat u$ is equivalent to the real-analytic function $\lambda \mapsto \ang{u(x), e^{-i\lambda \vdot x}}$ and $\abs{\hat u(\lambda)} \les \ang{\lambda}^N$ for some $N \geq 0$. 
We consider the complex extension of $\hat u$, and we call $\hat u(z)$ the \emph{Fourier-Laplace transform} of $u$. 
Note that
\[
\pdv{\hat u}{\bar z_i} = \ang{u, \pdv{\bar z_i} e^{-iz\vdot x}} = 0 \quad(i = 1, \dotsc, n), 
\]
so $\hat u(z)$ is complex-analytic in each component $z_i$.

We can also estimate the size of $\hat u(z)$:
\begin{lemma}
	If $u \in \EE'(\RR^n)$ and $\supp(u) \subseteq \clos{B_\delta}$, then there exist nonnegative constant $C, N$ such that
	\[
	\norm{\hat u(z)} \leq C (1 + \norm{z})^N e^{\delta \abs{\Im z}} \qquad \forall z \in \CC^n. 
	\]
\end{lemma}

\begin{proof}
	Let $\psi \in C^\infty(\RR)$ be such that $\psi(\tau) = 1$ for $\tau \geq -\frac12$ and $\psi(\tau) = 0$ for $\tau \leq -1$.  Introduce for $\eps > 0$
	\[
	\phi_\eps(x) \ceq \psi(\eps(\delta - \norm{x})),
	\]
	then we have $\phi_\eps \in \DD(\RR^n)$ with $\phi_\eps(x) = 0$ for $\norm{x}\geq \delta + \eps^{-1}$, and $\phi_\eps(x) = 1$ for $\norm{x} \leq \delta + \frac12\eps^{-1}$. In particular, every $\phi_\eps$ is 1 on a neighbourhood of $\supp(u) \subseteq \clos{B_\delta}$.

	Therefore, we have
	\[
	\hat u(z) = \ang{u(x), e^{-iz\vdot x}} = \ang{u(x), \phi_\eps(x) e^{-iz\vdot x}},
	\]
	and since $u \in \EE'(\RR^n)$, by the seminorm condition we have nonnegative $C, N$ such that
	\[
	\hat u(z) \leq C \sum_{\abs{\alpha} \leq N} \sup \abs{\pt^\alpha_x \qty(\phi_\eps(x) e^{-iz \vdot x})}.
	\]
	By definition we have $\abs{\pt_x^\beta \phi_\eps} \les \eps^{\abs{\beta}}$ while $\abs{\pt_x^\gamma e^{-iz \vdot x}} \les \norm{z}^{\abs{\gamma}} e^{(\delta + \eps^{-1}) \abs{\Im z}}$ on $\supp\phi_\eps$.

	Applying Leibniz, we obtain
	\[
	\abs{\hat u(z)} \leq C \sum_{\abs{\beta} + \abs{\gamma} \leq N} \norm{z}^{\abs{\beta}} e^{\abs{\gamma}} e^{(\delta + \eps^{-1}) \Im z}, 
	\]
	and this holds for all $\eps > 0$, so we can plug in $\eps = \norm{z}$ and obtain the result (since $\Im z / \norm{z}$ is bounded). 
\end{proof}

So if $u \in \EE'(\RR^n)$ with $\supp(u) \subseteq \clos{B_\delta}$, then $\hat u(z)$ is complex analytic and obeys $\abs{\hat u(z)} \leq (1 + \norm{z})^N e^{\delta \abs{\Im z}}$. The PWS theorem addresses the converse: if a complex analytic function obeys the estimate we just saw, is it the fourier transform of some distribution?

\begin{theorem}
	\begin{enumerate}[(a)]
		\item If $u \in \EE'(\RR^n)$ with $\ord(u) = N$ and $\supp u \subseteq \clos{B_\delta}$, then $\hat u(z)$ is entire and
		\begin{equation} \label{eq:fourier_transform_estimate}
			\abs{\hat u(z)} \les (1 + \norm{z})^N e^{\delta \abs{\Im z}}. 
		\end{equation}

	Conversely, if $U(z)$ is entire and satisfies \cref{eq:fourier_transform_estimate} for some $N$, then $U = \hat u$ for some $u \in \EE'(\RR^n)$ with $\supp(u) \subseteq \clos{B_\delta}$.

	\item If $u \in \DD(\RR^n)$ and $\supp(u) \subseteq \clos{B_\delta}$, then for $M = 0, 1, 2, \dotsc$ we have 
	\begin{equation}\label{eq:fourier_transform_estimate_2}
		\abs{\hat u(z)} \les_M (1 + \norm{z})^{-M} e^{\delta \abs{\Im z}}.
	\end{equation}
	Conversely, if $U(z)$ is entire and satisfies \cref{eq:fourier_transform_estimate_2}, then $U = \hat u$ for some $u \in \DD(\RR^n)$ with $\supp(u) \subseteq \clos{B_\delta}$. 
	\end{enumerate}
\end{theorem}

\begin{proof}
	\TODO write this out (lecture 12)
\end{proof}

\subsection{Oscillatory integrals}
We will study integrals of the form $\int_{\RR^n} e^{i\Phi(x, \theta)} a(x, \theta) \dd{\theta}$, where $\Phi$ is called the \emph{phase function} and $a$ the \emph{amplitude} of the signal. We wil use oscillations from the $e^{i\Phi}$-term to control the growth, while $a(\cdot, \theta)$ is allowed to grow modestly with $\theta$.

\begin{lemma}[Riemann-Lebesgue]
	If $f \in L^1(\RR)$, then $\abs{\hat f(\lambda)} \to 0$ when $\abs{\lambda} \to\infty$. 
\end{lemma}
\begin{proof}
	Assume $f \in L^1$ is continuous, then setting $x' = x + \pi/\lambda$, we have
	\begin{align*}
		\hat f(\lambda) &= \int e^{-i\lambda \cdot x} f(x) \dd{x} = \frac12\int \qty( e^{-i\lambda \vd x}f(x) \dd{x} + e^{-i\lambda \vd (x + \pi/\lambda)} f(x + \pi/\lambda))\dd{x} \\
		&= \frac12 \int e^{-i\lambda \vd x} \qty[ f(x) - f(x + \pi/\lambda)] \dd{x}
	\end{align*}
	Now let $\eps > 0$ and choose $R = R(\eps)$ such that $\int_{\norm{x}> R} \abs{f(x) - f(x + \pi/\lambda)} \dd{x} < \frac\eps2$.
	By the dominated convergence theorem, we can also choose $\lambda = \lambda(\eps, R)$ large enough such that
	\[
	\int_{\norm{x}< R} \abs{e^{-i\lambda x} \qty[f(x) - f(x + \pi/\lambda)]} \dd{x} < \frac\eps2. 
	\]  
	So for $f$ we have $\abs{\hat f(\lambda)} < \frac\eps2$ for $\abs{\lambda}$ large enough, which proves the result for continuous functions.

	Now we use that continuous functions are dense in $L^1$, so for any $f \in L^1$, pick $g \in C(\RR) \cap L^1$ such that $\norm{f - g}_{L^1} < \frac\eps2$, then it is easily checked that $\abs{\hat f(\lambda)} \leq \norm{f - g}_{L^1} + \abs{\hat g(\lambda)} < \eps $ for $\lambda$ sufficiently large, which proves the claim.

\end{proof}

Now suppose $\phi \in \DD(\RR)$ and $\Phi \in C^\infty(\RR)$ with $\Phi'$ nowhere 0. Consider
\[
I(\lambda) \ceq \int e^{i\lambda \Phi(\theta)} \phi(\theta) \dd{\theta}. 
\]
Note that $\abs{\Phi'(\theta)} \ges 1$ for $\theta \in \supp\phi$ (since $\supp\phi$ is compact), and we can write
\[
I(\lambda) = \int \frac1{i\lambda} \frac{\phi(\theta)}{\Phi'(\theta)} \dv{\theta}\qty(e^{i\lambda \Phi(\theta)}) \dd{\theta},
\]
and with repeated integration by parts we find that
\[
\abs{I(\lambda)} \leq \ang{\lambda}^{-N} \quad\text{for any $N \geq 0$}. 
\]
A natural question is what happens if $\Phi'(\theta) = 0$ somewhere.

\begin{lemma}[Stationary phase]
	Suppose $\Phi \in C^\infty(\RR)$ with $\Phi' \neq 0$ on $\RR \setminus \qty{0}$ and $\Phi(0) = \Phi'(0) = 0$, $\Phi''(0) \neq 0$. Then, for $\chi \in \DD(\RR)$, we have
	\[
	\int e^{i\lambda \Phi(\theta)} \chi(\theta) \dd{\theta} \les \ang{\lambda}^{-1/2}. 
	\]
\end{lemma}

\begin{proof}
	\TODO write this (lecture 13, a terrible proof with lots of differentiation and garbage). 
\end{proof}

### 2020-11-23 inverse_problems/C5.tex
wtfff

@@

	For an event $F \in \FF$, we define the \emph{probability of $F$ given $Y = y$} as $\EE[\ind_F \mid Y = y]$.
\end{definition}

@@

It can be shown that $(y, F) \mapsto \EE[\ind_F \mid Y= y]$ is a Markov kernel. This is clarified in the following theorem:

@@

We can use the above theorem to consider the density function of $U$ conditional on $Y$:
\begin{theorem}
	Under the assumptions of the previous lemma, we have $\PP(U \in \cdot \mid Y = y) \ll \nu_U$, and its $\nu_U$ density is
	\[
	g_{U \mid Y = y}(u) = \ind_{g_Y(y) > 0}  \frac{g_{U, Y}(u, y)}{g_Y(y)},
	\]
	which is uniquely defined up to null sets. 
\end{theorem}

>>>>

\begin{definition}
	Let $(\Omega, \FF, \mu)$ be a measure space and let $g \colon (\Omega, \FF) \to (\RR, \BB \RR)$ be measurable and non-negative. Then the map 
	\[
	\nu \colon \FF \to [0, \infty] \colon F \mapsto \int_F g \dd{\mu}
	\]
	defines a measure on $(\Omega, \FF)$, called the \emph{measure with ($\mu$-)density $g$}. If $\nu$ is a probability measure, $g$ is called a \emph{($\mu$-)probability density}. 
\end{definition}

@@

	%For an event $F \in \FF$, we define the \emph{probability of $F$ given $Y = y$} as $\EE[\ind_F \mid Y = y]$.
\end{definition}

@@

It can be shown that $(y, F) \mapsto \EE[\ind_F \mid Y= y]$ (which represents the probability of $F$ given $Y = y$) is a Markov kernel. This is clarified in the following theorem:

@@

We can use the above theorem to consider the density function of $U$ conditional on $Y$:
\begin{theorem}\label{thm:conditional_density}
	% TODO niet helemaal hetzelfde als thm 6.3.10 maar whatever
	Under the assumptions of the previous lemma, we have $\PP(U \in \cdot \mid Y = y) \ll \nu_U$, and its $\nu_U$ density is
	\[
	g_{U \mid Y = y}(u) = \ind_{g_Y(y) > 0}  \frac{g_{U, Y}(u, y)}{g_Y(y)},
	\]
	which is uniquely defined up to null sets. 
\end{theorem}

\begin{definition}
	Let $g_U, g_Y, g_{U, Y}, g_{U\mid Y= y}, g_{Y \mid U = u}$ be the probability densities from the previous theorem. Then $g_U$ is called the \emph{marginal probability density} of $U$, $g_{U, Y}$ is called the \emph{joint probability density} of $U$ and $Y$, and $g_{U \mid Y = y}$ is called the \emph{conditional density} of $U$ given $Y = y$.  
\end{definition}

\subsection{Bayesian statistics}
\subsubsection{Statistical models}
\begin{definition}
	Let $\XX, \YY$ be separable Banach spaces. We call $\XX$ the \emph{parameter space} and $\YY$ the \emph{data space}. Let $\Pp \ceq \qty{M(\cdot \mid u) : u \in \XX}$, where $M$ is a Markov kernel from $(\XX, \BB\XX)$ to $(\YY, \BB\YY)$. The tuple $(\YY, \Pp)$ is called the \emph{statistical model}. 
\end{definition}

Usually, we distinguish between \emph{parametric models} (where $\XX$ is finite-dimensional) and \emph{nonparametric models}. However, in Bayesian statistics, this distinction rarely matters.

Given a ``true'' parameter $u^* \in \XX$, a distribution $Y \sim M(\cdot \mid u^*)$ and a realisation $y$ of $Y$, we aim to find $u^*$ based on $y$. The probability measure $M(\cdot \mid u^*)$ is called the \emph{data-generating distribution}.

\begin{definition}
	Let $(\YY, \Pp)$ be a statistical model and $L \colon (\XX \times \YY, \BB\XX \otimes \BB\YY) \to (\RR, \BB\RR)$ such that
	\[
	\Pp = \qty{F \mapsto \int_F L(y \mid u) \dd\mu(y) : u \in \XX}
	\]
	for some measure $\mu$ on $(\YY, \BB\YY)$. Then $L$ is called the \emph{(data) likelihood}. 
\end{definition}
The likelihood is a conditional density $g_{Y\mid U = u}$ for some random variable $U$.

\subsubsection{Bayes' formula}
Assume we are given a likelihood $L = f_{Y \mid U = u}$ for some $U$. We are interested in finding $f_{U \mid Y = y}$, the posterior density function of $U$. For this we use Bayes's formula:
\begin{theorem}[Bayes]
	Let $U, Y$ be random variables as in \cref{thm:conditional_density}. Then we have
	\[
	g_{U \mid Y =y}(u) = \frac{g_{Y \mid U = u}(y) g_{U}(u)}{g_Y(y)}
	\]
	for $u \in \XX$, $\nu_U$-a.e., and $y \in \YY$, $\PP(Y \in \cdot)$-a.e.\ with $g_Y(y) > 0$. 
\end{theorem}

### 2020-11-23 modern_statistical_methods/C3.tex
wtfff

>>>>

\begin{definition}
	Let $S, U \in \RR^{n \times d}$ be matrices with orthogonal columns. We define the \emph{$\sint$ distance} between the column space of $S$ and $U$ as
	\[
	\sint(S,U) \ceq \norm{\Pi_S(I - \Pi_U)}_F = \norm{\Pi_U(I - \Pi_S)}_F
	\]
\end{definition}
To see that the last equality holds, note that
\begin{align*}
	\norm{\Pi_S(I - \Pi_U)}_F^2 &= \Tr\qty((I - \Pi_U) \Pi_S^2 (I - \Pi_U)) = \Tr\qty(\Pi_S(I - \Pi_U)) = \Tr(\Pi_S) - \Tr(\Pi_S\Pi_U) \\
	&= \rank(S) - \Tr(\Pi_S\Pi_U) = \rank(U) - \Tr(\Pi_U\Pi_S) = \norm{\Pi_U(I - \Pi_S)}_F^2. 
\end{align*}

In the fourth example sheet we prove the following:
\begin{proposition} \label{prop:sint_distance}
	Let $S, U$ be matrices with orthogonal columns and equal rank. Then
	\[
	\min_{O \in \RR^{d \times d} \text{ }\Rm{orthogonal}} \norm{S - UO}_F \leq \sqrt2 \sint(S, U). 
	\]
\end{proposition}

\begin{theorem}[Davis-Kahan]
	Let $\Sigma$ be a covariance matrix with eigenvector decomposition $\Sigma = U\Lambda U\T$ (where $\lambda_1 \geq \dotsb \geq \lambda_d$). Write $\Sigma = U_0 \Lambda_0 U_0\T + U_1 \Lambda_1 U_1\T$ where $U_0 = [u_1, \dotsc, u_r]$.

	For a perturbed matrix $\hat\Sigma = \Sigma + E$ with $E$ symmetric, define an analogous decomposition $\hat\Sigma = \hat U_0 \hat\Lambda_0 \hat U_0\T + \hat U_1 \hat\Lambda_1 \hat U_1\T$. If $\lambda_r - \hat\lambda_{r+1} > 0$, then
	\[
	\sint (U_0, \hat U_0)  \leq \frac{\norm{E}_F}{\lambda_r - \hat\lambda_{r+1}}. 
	\]
\end{theorem}

\begin{proof}
	We write
	\[
	\norm{\Pi_{U_0}(I - \Pi_{\hat U_0})}_F = \norm{U_0 U_0\T \hat U_1 \hat U_1\T}_F = \norm{U_0\T \hat U_1}_F, 
	\]
	and note that 
	\[
	\Sigma U_0 = U_0\Lambda U_0\T U_0 + U_1 \Lambda_1 U_1\T U_0 =  U_0\Lambda_0, \quad \hat\Sigma \hat U_1 = \hat U_1 \hat\Lambda_1. 
	\]
	We therefore find 
	\begin{align*}
	E U_0 &= (E + \Sigma) U_0 - \Sigma U_0 = \hat\Sigma U_0 - U_0 \Lambda_0 \\
	\hat U_1\T EU_0 &= \hat U_1\T \hat\Sigma U_0 - \hat U_1\T U_0 \Lambda_0 = \hat\Lambda_1 \hat U_1\T U_0 - \hat U_1\T U_0 \Lambda_0. 
	\end{align*}
By the triangle inequality we have
\begin{align*}
\norm{\hat U_1\T E U_0}_F &\geq \norm{\hat U_1\T U_0 \Lambda_0}_F - \norm{\hat\Lambda_1 \hat U_1\T U_0}_F \\
&\overset\star\geq \lambda_r \norm{\hat U_1\T U_0} - \hat\lambda_{r+1} \norm{\hat U_1\T U_0}_F = (\lambda_r - \hat\lambda_{r+1}) \norm{U_1\T U_0}_F = (\lambda_r - \hat\lambda_{r+1}) \sint(U_0, \hat U_0). 
\end{align*}
Noting that $\norm{\hat U_1\T E U_0}_F \leq \norm{E}_F$,
since
\[
\norm{E}_F = \norm{\hat U\T E U} = \norm{\mqty[\hat U_0\T \\ \hat U_1\T ] E \mqty[U_0 & U_1]}_F \geq \norm{\hat U_1\T E U_0}_F. 
\] yields the result. 
\end{proof}

The Davis-Kahan theorem is not directly useful since the result depends on the gap $\lambda_r - \hat\lambda_{r+1}$. However, the following variant was proved in 2014, whose result depends only on the eigenvalues of $\Sigma$:
\begin{theorem}
	If $\lambda_r - \lambda_{r+1} > 0$, then
	\[
	\sint (U_0, \hat U_0) \leq \frac{2\min \qty{\sqrt r \norm{E}_\Rm{op}, \norm{E}_F}}{\abs{\lambda_r - \lambda_{r+1}}}. 
	\]
\end{theorem}

\subsubsection{The spiked covariance model}
As an application of the Davis-Kahan theorem, consider the model where
\[
\Sigma = I + \theta v v\T, \qquad  v\in S^{d-1}. 
\]
Any random vector with covariance $\Sigma$ can be represented as $\eta \sqrt\theta v + \eps$, where $\eta, \eps_1, \dotsc, \eps_d$ are independent, mean 0, variance 1 random variables. The samples are a random ``spike'' with scale $\sqrt\theta$ in a fixed direction $v$, with independent noise added to each entry.

The eigenvalues of $\Sigma$ are $1+\theta, 1, \dotsc, 1$, and $v$ is the top eigenvector. Using the variant of the Davis-Kahan theorem and \cref{prop:sint_distance} we find 
\[
\min_{z = \pm 1} \norm{v - z\hat v} \leq \sqrt2 \sint(v, \hat v) \leq \frac{\norm{\hat \Sigma - \Sigma}_\Rm{op}}{(1 + \theta) - 1}
\]
If we assume that the rows of $X$ are vector sub-Gaussian, we can bound $\norm{\hat\Sigma - \Sigma}_\Rm{op}$ in probability, and we find that 
\[
\PP\qty(\min_{z = \pm 1} \norm{v - z\hat v} \geq C \frac{1 + \theta}{\theta} \qty(\frac{d + \delta}{n} \lor \sqrt{\frac{d + \delta}{n}})) \leq e^{-\delta}. 
\]
We conclude that if $\frac{1+\theta}{\theta} \sqrt{\frac dn} \to 0$, then $\hat v$ estimates $v$ consistently.

\subsubsection{High-dimensional asymptotics}
{\color{red} Non-examinable}.

\subsection{Sparse PCA}
We have seen that in the spiked covariance model, PCA is consistent when $d/n \to 0$, but PCA can fail if $d/n$ is bounded away from zero. Considering the case $d/n \to \gamma > 0$, we can ask if there exists a better estimator. Minimax lower bounds show that in general, the answer is negative, and therefore we must place some assumptions on $\Sigma$.

Again, we assume $\Sigma = \theta vv\T + I$. Define
\[
\norm{v}_0 \ceq \abs{\qty{k \mid v_k \neq 0}} \quad\text{(the amount of nonzero components of $v$)}
\]
We will assume that $v$ is spare, i.e., $\norm{v}_0 = k \ll d$. This means that the extra variance in the data along the ``spike'' only involved a few variables. 
We can now define an estimator similar to best subset selection
\[
\hat v^{(0, k')} \ceq \argmax_{w \in S^{d-1}, \norm{w}_0 \leq k'} w\T \hat\Sigma w. 
\]

\begin{proposition}
	Let $\hat\Sigma$ be the sample covariance of $n$ i.i.d.\ $N(0, \Sigma)$ random variables. Suppose $k \leq d/2$ and $k \leq k'$, then 
	\[
	\PP\qty(\min_{z = \pm 1} \norm{v - z\hat v^{(0, k')}}_2 \geq  C\frac{1+\theta}{\theta} \qty[ \sqrt{\eta_n \lor \eta_n}]) \leq e^{-\delta} \quad\forall \delta > 0
	\]
	where 
	\[
	\eta_n = \frac{(k + k') \log(de^2/(k+ k')) + \delta}{n}.
	\]
\end{proposition}
Note that the dependence on $d$ is now logarithmic, so morally, we only need $\frac{\max(k, k')}{n} \to 0$ for consistency.

The problem with this estimator is that it is hard to compute. The SCoTLASS estimator relaxes the estimator with 
\[
\hat v^{(1, \lambda)} \ceq \argmax_{w \in S^{d-1}, \norm{w}_1 \leq \lambda} w\T \hat\Sigma w.
\]
The optimisation problem is not convex (since we are maximising a convex function instead of minimising), but is easier to compute and has similar convergence rates as the estimator $\hat v^{(0, k)}$. We can relax the problem even further (see example sheet) which gives an easily computable solution.

It has been shown that sparce PCA has no estimators computable in polynomial time which achieve minimax rates for certain distribution classes, which means that sparce PCA has some inherent computational difficulty.
